---
title: analyze flow cytometry data from 2020-12-10
author: Bin He
date: 2020-12-14
output: 
  html_notebook:
    toc: true
    toc_float: true
    df_print: default
---

```{r setup}
require(tidyverse)
require(cowplot)
```

## Goal

Testing the brightness of Pho4-GFP (mNeon and EGFP) either on a plasmid or integrated at the leu2 or endogenous Pho4 locus. also evaluate the population heterogeneity and use the PHO5pr-mCherry reporter to evaluate the functionality of the Pho4 variants

## Material and methods

### table of strains
```{r echo = F}
samples <- read_tsv("strains-genotype.tsv", col_types = "ccccccccD")
(samples)
```

### methods
See gDoc for details. Some strains didn't grow to saturation after the first overnight culture and thus the o/n culture instead of the second growth, which was at mid log phase at the time of harvest, was used for flow cytometry.

## Data and analysis
```{r}
dat <- read_csv("20201210-testing-Pho4-GFP-on-plasmid-or-integrated-with-PHO5pr-reporter.csv")
dat <- dat %>% mutate(Pho4GFP = factor(Pho4GFP, levels = c("No-GFP","ScPho4-mNeon","ScPho4-EGFP", "CgPho4-mNeon")))
```

### Pho4-GFP expression levels
First look at the level of Pho4-GFP compared with the negative controls.
```{r, echo = F, fig.height = 7, fig.width = 10}
gfp.bg <- dat %>% filter(Pho4GFP == "No-GFP", Parameter == "Pho4-GFP") %>% pull(Median) %>% mean()
p1 <- dat %>% filter(Parameter == "Pho4-GFP", Median < 3000) %>% 
  ggplot(aes(x = Sample, y = Median, fill =  Group)) + 
  geom_bar(stat = "identity", position = position_dodge2(width = 0.9)) +
  geom_text(aes(label = ifelse(Growth == "ovn", "*", ""), hjust = -0.2, vjust = 0.8), 
            size = 5) +
  geom_hline(yintercept = gfp.bg, linetype = "dashed", alpha = 0.6) +
  facet_grid(Pho4GFP~., scales = "free_y", space = "free_y") +
  scale_fill_brewer(type = "div") + ylab("MFI") +
  coord_flip() + scale_y_continuous(limits = c(0,3000), expand = c(0.01,0)) +
  theme_bw() + theme(legend.position = "none")

# include the outliers
p2 <- dat %>% filter(Parameter == "Pho4-GFP") %>% 
  ggplot(aes(x = Sample, y = Median, fill =  Group)) + 
  geom_bar(stat = "identity", position = position_dodge2(width = 0.9)) +
  geom_text(aes(label = ifelse(Growth == "ovn", "*", ""), hjust = -0.2, vjust = 0.8), 
            size = 5) +
  geom_hline(yintercept = gfp.bg, linetype = "dashed", alpha = 0.6) +
  facet_grid(Pho4GFP~., scales = "free_y", space = "free_y") +
  scale_fill_brewer(type = "div") + ylab("MFI") + xlab("") +
  coord_flip() + scale_y_continuous(expand = c(0.01,0)) + 
  theme_bw()

# code below from https://wilkelab.org/cowplot/articles/plot_grid.html
# plotting both
plot_row <- plot_grid(p1, p2, rel_widths = c(1,1.3), labels = "AUTO")

# now add the title
title <- ggdraw() +
  draw_label(
    "GFP fluorescent intensity (arbitrary units), dashed line indicates background fluorescence",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )
plot_grid(
  title, plot_row,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.07, 1)
)
```
**Figure 1 Pho4-GFP intensity by strain.** Pho4-GFP intensities were quantified on an Attune NxT flow cytometer using 400 mV on the BL1 channel. Acquisition rate is 200 ul/min. Strain IDs are listed on the vertical axis while the Median Fluorescent Instensity (MFI, arbitrary units) on the horizontal axis. An asterisk next to a bar indicates an overnight growth instead of a mid-log phase growth was used for quantification. **(A)** omits A3, A5 and A6, which appear to be outliers to the rest. C2, B3 and D1 also appear to be much brighter than the rest. **(B)** includes all samples. Notice how the scale is dominated by A3,5,6.

_Discussion_

- With limited samples for the endogenously integrated Pho4, it appears that this method generates less clone-to-clone variability.
- Comparing G and H, which are endogenously integrated with mNeon or EGFP tag, it is clear that the former is brighter. The same comparison with the exogenously integrated strains, e.g. A vs B or D vs E, gave less clear answers. Specifically, B1-3 appear to show higher intensity than A's, although a couple of A clones, i.e. A3, A5 and A6 are outliers as listed above. Also, in my previous test on 2020-11-09, B clones are generally dimmer than A clones. So there may be some reproducibility issues.
- The outliers mentioned above are all within the exogenous group. My suspicion is that the ends-in integration method by the pRS305 based plasmids are likely to result in multiple integrations, which could happen at higher rate when a larger amount of digested plasmid is used in the transformation.
- The plasmid-borne Pho4-GFP strains didn't grow to saturation as the others did, likely due to expression levels of the Leu2 gene from the plasmids. As a result, for most of them I used the o/n culture. The two mid-log phase samples, M1 and M2, appear to be brighter than the o/n samples, although one needs to repeat this to know for sure.

Next let's see the population heterogeneity within each strain, using the percent robust CV measure offered by the Attune software.
```{r plot_robust_CV, fig.height = 6, fig.width = 6}
# let's also check the "spread" using %CV or %robustCV to see which methods produce more heterogeneity among cells
dat %>% filter(Parameter == "Pho4-GFP") %>% 
  ggplot(aes(x = Sample, y = rCV, fill =  Group)) + 
  geom_bar(stat = "identity", position = position_dodge2(width = 0.9)) +
  geom_text(aes(label = ifelse(Growth == "ovn", "*", ""), hjust = 1.2, vjust = 0.8), 
            size = 5, color = "white") +
  facet_grid(Pho4GFP~., scales = "free_y", space = "free_y") +
  scale_fill_brewer(type = "div") + ylab("%rCV") +
  coord_flip() + scale_y_continuous(expand = c(0.01,0)) +
  labs(title = "%rCV for Pho4-GFP fluorescent intensity", subtitle = "Asterisks indicate o/n instead of mid-log phase culture") +
  theme_bw()
```
**Figure 2 Pho4-GFP fluorescence cell-to-cell variability within each strain.** Pho4-GFP intensities were quantified on an Attune NxT flow cytometer using 400 mV on the BL1 channel. Acquisition rate is 200 ul/min. Strain IDs are listed on the vertical axis while the percent robust CV (%rCV) on the horizontal axis. An asterisk next to a bar indicates an overnight growth instead of a mid-log phase growth was used for quantification.

_Discussion_

- P1-4 had the highest level of spread, although it is unclear if that's due to the o/n growth or due to the strain itself.
- Note that the outliers in GFP fluorescence in **Fig. 1**, which are exogenously integrated, _don't_ have higher variability, suggesting that they are stable, but perhaps multiple integrants.

### PHO5pr-mCherry reporter levels
```{r, fig.height = 7, fig.width = 10}
rfp.bg <- mean(dat %>% filter(!PHO5RFP, Parameter == "PHO5pr-mCherry") %>% pull(Median))
rfp.basal <- dat %>% filter(PHO5RFP, Pho4 == "pho4-", Parameter == "PHO5pr-mCherry") %>% pull(Median) %>% mean()

# removed a few outliers with MFI > 20k
p1 <- dat %>% filter(PHO5RFP, Parameter == "PHO5pr-mCherry", Median < 20000) %>% 
  ggplot(aes(x = Sample, y = Median, fill =  Pho4GFP, color = PHO2)) + 
  geom_bar(stat = "identity", position = position_dodge2(width = 0.9)) +
  geom_hline(yintercept = rfp.basal, linetype = 2, alpha = 0.6) +
  facet_grid(Pho4 ~ ., scales = "free_y", space = "free_y") + ylab("MFI") +
  scale_fill_brewer(type = "div") + scale_color_manual(values = c("red", NA)) +
  coord_flip() + scale_y_continuous(limits = c(0, 15000), expand = c(0.01,0.02)) +
  theme_bw() + theme(legend.position = "none")

# all data
p2 <- dat %>% filter(PHO5RFP, Parameter == "PHO5pr-mCherry") %>% 
  ggplot(aes(x = Sample, y = Median, fill =  Pho4GFP, color = PHO2)) + 
  geom_bar(stat = "identity", position = position_dodge2(width = 0.9)) +
  geom_hline(yintercept = rfp.basal, linetype = 2, alpha = 0.6) +
  facet_grid(Pho4 ~ ., scales = "free", space = "free") + ylab("MFI") + xlab("") +
  scale_fill_brewer(type = "div") + scale_color_manual(values = c("red", NA)) +
  coord_flip() + scale_y_continuous(expand = c(0.01,0.05)) +
  theme_bw()

# plotting both
plot_grid(p1, p2, rel_widths = c(1,1.35), labels = "AUTO")
```
**Figure 3 PHO5pr-mCherry reporter fluorescence intensity strain.** PHO5pr-mCherry intensities were quantified on an Attune NxT flow cytometer using 372 (?) mV on the YL2 channel. Acquisition rate is 200 ul/min. Strain IDs are listed on the vertical axis and the  Median Fluorescent Intensity (MFI, arbitury units) on the horizontal axis. An asterisk next to a bar indicates an overnight growth instead of a mid-log phase growth was used for quantification. The background fluorescence is just 85 units, i.e. very little background in the YL2 channel. The dashed vertical line indicates the basal level expression from the reporter without Pho4 (average of yH295, yH296, both of which lack Pho4 but has the reporter). **(A)** omits G2, G3 and G4, while **(B)** includes all samples.

_Discussion_

- CgPho4 clearly has no dependence on Pho2 when assayed in _S. cerevisiae_, largely consistent with my previous results (although my RNA-seq has indicated a modest dependence with ~30-40% reduction in mRNA levels of _PHO5_ driven by CgPho4 without Pho2.)
- ScPho4 has strong dependence on Pho2. P1-4 and B1,2,4 both showed that without Pho2, ScPho4 doesn't induce reporter expression on its own. Some anormalies exist however, as will be discussed below.
- H1-3 are supposed to be ScPho4-EGFP with the PHO5pr-mCherry and also with ScPho2. However, the results here suggest that this strain had little reporter induction. It may be that the integrated ScPho4-EGFP, even though it is expressed, is not functional. Is this consistent with Lindsey's results, including flow cytometry and liquid phosphatase assays?
- G2, 3, 4 have much higher levels than everyone else. Given that the G strains were constructed by Lindsey by integrating the pRS306-PHO5pr-mCherry-URA3 construct into the ura3 locus using the ends-in transformation, I wonder if G2,3 and 4, just like those leu2 integrated Pho4 variants, represent multiple integrations.
- Lastly, note that the several A clones with high Pho4-GFP (A3, A5 and A6) also showed higher levels of reporter expression, suggesting that increasing the dose of Pho4 can certainly partially compensate for the loss of Pho2.


## Conclusions
- mNeon appears to be as good as, if not better than EGFP in terms of both brightness and keeping the tagged Pho4 functional.
- the ends-in integration seems to generate multiple integrations frequently (maybe a result of excess digested plasmid? but we are also fighting relatively low transformation efficiency)
- increasing the dose of ScPho4 can partially compensate for the loss of Pho2, which further highlights the importance of ensuring equal or highly similar levels of expression for all constructs to be tested.