---
title: "select-experiments"
author: "Bin He"
date: "2023-01-10 (updated `r Sys.Date()`)"
output: html_notebook
---

```{r}
require(tidyverse)
require(cowplot)
```

## 2023-01-16, CY, Sc/Cg

Chris performed qPCR to quantify candidate gene induction in the time course under three [Pi] in both _S. cerevisiae_ and _C. glabrata_

```{r}
raw <- read_tsv("20230116-CY-PHO84-SPL2-time-course-raw.tsv", col_types = cols(), comment = "#")
dat <- select(raw, Sample, Pi, Minute, Replicate, Target, Task, Cq) %>% 
  mutate(Pi = factor(Pi, levels = c(7, 0.06, 0)),
         Minute = factor(Minute, levels = c(0, 2, 45, 120)),
         Target = factor(Target, levels = c("scact1", "Scpho84", "Scspl2"),
                         labels = c("ScACT1", "ScPHO84", "ScSPL2")),
         Flag = ifelse(Cq >= 30, "Fail", "Pass"))
dCt <- dat %>% 
  filter(Task == "UNKNOWN") %>% 
  group_by(Pi, Minute, Replicate, Target) %>% 
  summarize(mCq = mean(Cq), .groups = "drop_last") %>% # drop grouping by Target
  mutate(dCt = mCq - mCq[Target == "ScACT1"]) %>% 
  filter(Target != "ScACT1")

ref <- ungroup(dCt) %>% filter(Minute == 0) %>% select(-Pi, - Minute, -mCq)
ddCt <- filter(dCt, Minute != 0) %>% 
  left_join(ref, by = c("Replicate", "Target"), suffix = c("", ".ref")) %>% 
  mutate(ddCt = dCt.ref - dCt)
```

```{r}
p <- ggplot(ddCt, aes(x = Pi, y = ddCt)) +
  geom_bar(aes(fill = Minute), stat = "summary", fun = "mean", 
           position = position_dodge(.8), width = 0.5, color = "black") +
  geom_point(aes(group = Minute), size = 1,
             position = position_dodge(0.8)) +
  scale_fill_grey(start = 0.5) +
  facet_wrap(~Target) + xlab("Pi (mM)") +
  theme_cowplot() + theme(strip.text = element_text(face = 3))
p
ggsave("20230119-CY-qPCR-time-course-plot.png", width = 5, height = 3)
```

**_Conclusion_**

- ScPHO84 was clearly and strongly induced by 120 min, which is what we expected.
- We expected stronger induction (~30 fold) of ScSPL2 by 30 min based on Zhou & O'Shea 2011
- Looking back at Gurvich et al 2017 data, we were surprised to see that ScPHO84 was shown to be at most 2^3 = 8 fold induced throughout the time course at 0 mM Pi. We wondered if it was due to their unique sequencing and analysis procedures.
