---
title: "Determine stress conditions and timepoints for genomics profiling"
author: Bin He
date: 2022-03-07
output: html_notebook
---

```{r}
require(tidyverse)
require(cowplot)
```

## Goal

Summarize qRT-PCR and growth curve results performed by VKS from Nov 2021 to Jan 2022 in order to determine the time points and [H2O2] for the time course RNAseq.

## Materials and Methods

### qRT-PCR
| **Date**  | **Species**   | **[H2O2] mM**  | **Time points (min)**  | **Targets**      |
|-----------|---------------|----------------|------------------------|------------------|
| 2022.01   | _S. cerevisiae_ | 0, 0.2, 0.5, 5 | 0, 10, 20, 40, 120     | CTT1, GRE2       |
| 2021.12   | _C. glabrata_   | 0, 5, 10, 100  | 0, 10, 20, 40, 80, 100 | CTA1, GRE2       |
| 2021.11   | _C. glabrata_   | 0, 5           | 40                     | CTA1, GRE2, TRX2 |

```{r}
dat <- read_tsv("../input/20220306-summary-H2O2-time-course-gene-induction.tsv")
```

### growth curve
| **Date**  | **Species**   | **[H2O2] mM**  | **Time points (min)**  | **Targets**      |
|-----------|---------------|----------------|------------------------|------------------|
| 2022.01   | _S. cerevisiae_ | 0, 0.2, 0.5, 5 | 0, 10, 20, 40, 120     | CTT1, GRE2       |

## Plot
```{r}
dat1 <- dat %>% 
  mutate(FC = 2^log2FC,
         time = factor(treatTime),
         H2O2 = ordered(h2o2, levels = paste(c(0.2, 0.5, 1, 5, 10, 100), "mM")))

dat1 %>% filter(gene %in% c("CTA1", "CTT1", "GRE2")) %>% 
  ggplot(aes(x = time, y = FC, color = H2O2, group = H2O2)) +
  geom_point() + stat_summary(geom = "line", fun.y = mean) +
  scale_color_manual(values = c(
    "0.2 mM" = "yellow3",
    "0.5 mM" = "orange2",
    "1 mM" = "darkred",
    "5 mM" = "limegreen",
    "10 mM" = "darkgreen",
    "100 mM" = "purple4"
  )) +
  facet_wrap(~ paste(species, gene, sep = " - ")) +
  ylab("Fold induction") + xlab("Time (min)") +
  theme_cowplot() + panel_border() + background_grid(major = "y") +
  theme(strip.text = element_text(face = 3))
```

```{r}
ggsave("../output/20220307-qRT-PCR-Sc-Cg-time-course.png", width = 7, height = 5, bg = "white")
```

Plot the two species separately
```{r}
p1 <- dat1 %>% 
  filter(gene %in% c("CTA1", "CTT1", "GRE2"), species == "S. cerevisiae") %>% 
  ggplot(aes(x = time, y = FC, color = H2O2, group = H2O2)) +
  geom_point() + stat_summary(geom = "line", fun = mean, size = 1) +
  scale_color_viridis_d(begin = 0.2) + #scale_y_sqrt() +
  facet_wrap(~ gene) +
  ylab("Fold induction") + xlab("Time (min)") +
  theme_cowplot(font_size = 16) + panel_border() + background_grid(major = "y") +
  theme(strip.text = element_text(face = 3))
p2 <- dat1 %>% 
  filter(gene %in% c("CTA1", "CTT1", "GRE2"), species == "C. glabrata") %>% 
  ggplot(aes(x = time, y = FC, color = H2O2, group = H2O2)) +
  geom_point() + stat_summary(geom = "line", fun = mean, size = 1) +
  scale_color_viridis_d(begin = 0.2) + #scale_y_sqrt() +
  facet_wrap(~ gene) +
  ylab("Fold induction") + xlab("Time (min)") +
  theme_cowplot(font_size = 16) + panel_border() + background_grid(major = "y") +
  theme(strip.text = element_text(face = 3))
ggsave("../output/20220419-qRT-PCR-Sc-time-course.png", p1, width = 7, height = 4)
ggsave("../output/20220419-qRT-PCR-Cg-time-course.png", p2, width = 7, height = 4)
```

